PyBOP: A Python package for battery model optimisation and parameterisation

### 2024-12-18 joss/paper.preprint.tex
update and vectorise figures.

\usepackage[affil-it]{authblk}
\usepackage{orcidlink}
%% \renewcommand\Authsep{, }
\setlength{\affilsep}{1em}
\author[1%
  %
  \ensuremath\mathparagraph]{Brady Planden%
    \,\orcidlink{0000-0002-1082-9125}\,%
    }
\author[1,2%
  %
  ]{Nicola E. Courtier%
    \,\orcidlink{0000-0002-5714-1096}\,%
    }
\author[3%
  %
  ]{Martin Robinson%
    \,\orcidlink{0000-0002-1572-6782}\,%
    }
\author[4%
  %
  ]{Agriya Khetarpal%
    \,\orcidlink{0000-0002-1112-1786}\,%
    }
\author[2,5%
  %
  ]{Ferran Brosa Planella%
    \,\orcidlink{0000-0001-6363-2812}\,%
    }
\author[1,2%
  %
  ]{David A. Howey%
    \,\orcidlink{0000-0002-0620-3955}\,%
    }

@@

The Python Battery Optimisation and Parameterisation (\texttt{PyBOP})
package provides methods for estimating and optimising battery model
parameters, offering both deterministic and stochastic approaches with
example workflows to assist users.

\texttt{PyBOP} enables parameter identification from data for various
battery models, including both the electrochemical and equivalent
circuit models provided by the popular open-source \texttt{PyBaMM}
package (\citeproc{ref-Sulzer:2021}{Sulzer et al., 2021}). Using the
same workflow, \texttt{PyBOP} can also be used for design optimisation
under user-defined operating conditions across a variety of model
structures and design goals. \texttt{PyBOP} facilitates optimisation
using a range of methods, with diagnostics for examining optimiser
performance and convergence of the cost and corresponding parameters.
Identified parameters can be used for prediction, on-line control, and
design optimisation to accelerate research and development and improve
battery utilisation.

@@

\texttt{PyBOP} is a Python package that provides a user-friendly,
object-oriented interface for optimising battery model parameters.
\texttt{PyBOP} leverages the open-source \texttt{PyBaMM} package
(\citeproc{ref-Sulzer:2021}{Sulzer et al., 2021}) to formulate and solve
battery models. Together, these packages serve a broad audience
including students, engineers, and researchers in academia and industry,
enabling the use of advanced models where previously this was not
possible without specialised knowledge of battery modelling, parameter
inference and software development. \texttt{PyBOP} emphasises clear and
informative diagnostics and workflows to support users with varying
levels of domain expertise, and provides access to a wide range of
optimisation and sampling algorithms. These methods are provided through
interfaces to \texttt{PINTS} (\citeproc{ref-Clerx:2019}{Clerx et al.,
2019}), \texttt{SciPy} (\citeproc{ref-SciPy:2020}{Virtanen et al.,
2020}), and \texttt{PyBOP}'s own implementations of algorithms such as
adaptive moment estimation with weight decay (AdamW), gradient descent,
and Cuckoo search.

\texttt{PyBOP} supports the battery parameter exchange (BPX) standard
(\citeproc{ref-BPX:2023}{Korotkin et al., 2023}) for sharing battery
parameter sets. These parameter sets are typically costly to obtain due
to the specialised equipment and time required for characterisation
experiments, the need for battery domain knowledge, and the
computational cost of parameter estimation. \texttt{PyBOP} reduces the
requirements for the latter two categories by providing fast parameter
estimation, standardised workflows, and parameter set interoperability
(via BPX).

This package complements other lithium-ion battery modelling packages
built around \texttt{PyBaMM}, such as \texttt{liionpack} for battery
pack simulation (\citeproc{ref-Tranter2022}{Tranter et al., 2022}) and
\texttt{pybamm-eis} for fast numerical computation of the
electrochemical impedance for any battery model, since the identified
parameters from \texttt{PyBOP} are easily exportable to these other
packages.

@@

\texttt{PyBOP} has a layered structure designed to package up the
necessary functionality to compute forward predictions, process the
results, and run the optimisation and sampling algorithms. The forward
model is solved using the popular battery modelling package,
\texttt{PyBaMM}, with construction, parameterisation, and discretisation
managed by \texttt{PyBOP}'s model interface to \texttt{PyBaMM}. This
provides a robust object construction process with a consistent
interface between models and optimisers. The statistical methods and
optimisation algorithms are constructed to interface cleanly with the
forward model predictions. Furthermore, identifiability metrics are
provided with the estimated parameters (through Hessian approximation of
the cost functions in frequentist workflows, and posterior distributions
in Bayesian workflows).

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth,height=\textheight]{figures/PyBOP-high-level.pdf}
\caption{PyBOP's interface to supporting funding agencies, alongside a
visualisation of the general workflow for parameterisation and
optimisation. \label{fig:high-level}}
\end{figure}

\texttt{PyBOP} formulates the inference process into four key classes,
namely, the model, the problem, the cost/likelihood, and the
optimiser/sampler, as shown in \autoref{fig:classes}. Each of these
objects represents a base class with child classes constructing
specialised functionality for different workflows. The model class
constructs a \texttt{PyBaMM} forward model for a given set of model
equations, initial conditions, spatial discretisation, and numerical
solver. By composing \texttt{PyBaMM} directly into \texttt{PyBOP},
specialised models can be constructed alongside the standard models,
which can be modified and purposefully constructed for different
inference tasks. One such example is spatial re-discretisation, which is
required when one or more geometric parameters are being optimised. In
this situation, \texttt{PyBOP} rebuilds the \texttt{PyBaMM} model a
minimum number of times while maintaining the problem, cost, and
optimiser objects, providing improved performance benefits to users.
Alongside construction of the forward model, \texttt{PyBOP}'s model
class provides methods for obtaining sensitivities from the prediction,
enabling gradient-based optimisation algorithms. A forward prediction,
along with its corresponding sensitivities, is provided to the problem
class for processing and exception control. A standardised data
structure is then provided to the cost classes, which then computes a
distance, design, or likelihood-based metric for optimisation. For
deterministic optimisation, the optimisers minimise the cost function or
the negative log-likelihood if a likelihood class is provided. Bayesian
inference is provided by sampler classes, which accept the LogPosterior
class and sample from it using \texttt{PINTS} based Monte Carlo
algorithms at the time of submission. In the typical workflow, the
classes in \autoref{fig:classes} are constructed in sequence.

In addition to the core architecture, \texttt{PyBOP} provides several
specialised inference and optimisation processes. One such instance is
parameter inference from electrochemical impedance spectroscopy (EIS)
simulations. PyBOP discretises the EIS forward model into a sparse mass
matrix form with accompanying auto-differentiated Jacobian. These
objects are then translated into the frequency domain with a linear
solution used to compute the battery model impedance. In this situation,
the forward models are constructed within the spatial re-discretisation
workflow, allowing for geometric parameter inference from EIS
simulations. Furthermore, \texttt{PyBOP} builds on the \texttt{JAX}
(\citeproc{ref-jax:2018}{Bradbury et al., 2018}) numerical solvers
provided by \texttt{PyBaMM} by providing \texttt{JAX}-based cost
functions for automatic forward model differentiation with respect to
the parameters. This functionality provides a performance improvement,
and allows users to use many other JAX-based inference packages in order
to optimise their cost function, such as \texttt{Numpyro}
(\citeproc{ref-numpyro:2019}{Phan et al., 2019}), \texttt{BlackJAX}
(\citeproc{ref-blackjax:2024}{Cabezas et al., 2024}), and \texttt{Optax}
(\citeproc{ref-optax:2020}{DeepMind et al., 2020}).

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth,height=\textheight]{figures/PyBOP_components.drawio.png}
\caption{The core \texttt{PyBOP} architecture, showcasing the base class
interfaces. Each class provides a direct mapping to a classical step in
the optimisation workflow. \label{fig:classes}}
\end{figure}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4458}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2048}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3494}}@{}}
\caption{List of available model, problem and cost (or likelihood)
classes. \label{tab:subclasses}}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Battery Models
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Problem Types
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cost / Likelihood Functions
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Battery Models
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Problem Types
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cost / Likelihood Functions
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Single particle model (SPM) & Fitting problem & Sum squared error \\
SPM with electrolyte (SPMe) & Design problem & Root mean squared
error \\
Doyle-Fuller-Newman (DFN) & Observer & Minkowski \\
Many particle model (MPM) & & Sum of power \\
Multi-species multi-reaction (MSMR) & & Gaussian log likelihood \\
Weppner Huggins & & Maximum a posteriori \\
Equivalent circuit model (ECM) & & Volumetric energy density \\
& & Gravimetric energy density \\
& & Unscented Kalman filter \\
\end{longtable}

Similarly, the current algorithms available for optimisation tasks are
presented in \autoref{tab:optimisers}. It should be noted that
\texttt{SciPy} minimize comprises a range of gradient and non-gradient
methods. From now on, the point-based parameterisation and design
optimisation tasks will simply be referred to as optimisation tasks.
This simplification can be justified by examining
\autoref{eqn:parameterisation} and \autoref{eqn:design} and confirming
that deterministic parameterisation can be viewed as an optimisation
task to minimise a distance-based cost function.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4359}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2308}}@{}}
\caption{The currently supported optimisation algorithms classified by
candidate solution type, including gradient information.
\label{tab:optimisers}}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gradient-based
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Evolutionary
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
(Meta)heuristic
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gradient-based
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Evolutionary
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
(Meta)heuristic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Weight decayed adaptive moment estimation (AdamW) & Covariance matrix
adaptation (CMA-ES) & Particle swarm (PSO) \\
Improved resilient backpropagation (iRProp-) & Exponential natural
(xNES) & Nelder-Mead \\
Gradient descent & Separable natural (sNES) & Cuckoo search \\
SciPy minimize & SciPy differential evolution & \\
\end{longtable}

As discussed above, in addition to point-based optimisation estimates
\autoref{tab:subclasses}, \texttt{PyBOP} also provides Monte Carlo
sampling routines in order to recover distributions of likely parameter
values. The Monte Carlo methods construct a posterior distribution on
the inference parameters, which can be used to assess uncertainty and
practical identifiability. The individual sampler classes are currently
composed within \texttt{PyBOP} from the \texttt{PINTS} library, with a
base sampler class implemented for interoperability and direct
integration with the \texttt{PyBOP} model, problem, and likelihood
classes. The currently supported samplers are listed in
\autoref{tab:samplers}.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1624}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2393}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1368}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2051}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2564}}@{}}
\caption{Sampling methods supported by \texttt{PyBOP}, classified
according to the proposed method. \label{tab:samplers}}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gradient-based
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Adaptive
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Slicing
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Evolutionary
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Other
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gradient-based
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Adaptive
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Slicing
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Evolutionary
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Other
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Monomial Gamma & Delayed Rejection Adaptive & Rank---Shrinking &
Differential Evolution & Metropolis Random Walk \\
No-U-Turn & Haario Bardenet & Doubling & & Emcee Hammer \\
Hamiltonian & Haario & Stepout & & Metropolis Adjusted Langevin \\
Relativistic & Rao Blackwell & & & \\
\end{longtable}

@@

In general, battery models (after spatial discretisation) can be written
in the form of a differential-algebraic system of equations:
\begin{equation}
\frac{\mathrm{d} \mathbf{x}}{\mathrm{d} t} = f(t,\mathbf{x},\mathbf{\theta}),
\label{dynamics}
\end{equation} \begin{equation}
0 = g(t, \mathbf{x}, \mathbf{\theta})
\label{algebraic}
\end{equation} \begin{equation}
\mathbf{y}(t) = h(t, \mathbf{x}, \mathbf{\theta}),
\label{output}
\end{equation} with initial conditions \begin{equation}
\mathbf{x}(0) = \mathbf{x}_0(\mathbf{\theta}).
\label{initial_conditions}
\end{equation}

@@

Common battery models include various types of equivalent circuit models
(e.g.~the Thévenin model), the Doyle--Fuller--Newman (DFN) model
(\citeproc{ref-Doyle:1993}{Doyle et al., 1993};
\citeproc{ref-Fuller:1994}{Fuller et al., 1994}) based on porous
electrode theory and its reduced-order variants including the single
particle model (SPM) (\citeproc{ref-Planella:2022}{Brosa Planella et
al., 2022}), and the multi-species, multi-reaction (MSMR) model
(\citeproc{ref-Verbrugge:2017}{Verbrugge et al., 2017}). Simplified
models that retain acceptable predictive capabilities at a lower
computational cost are widely used, for example in battery management
systems, while physics-based models are required to understand the
impact of physical parameters on battery performance. This separation of
complexity traditionally results in multiple parameterisations for a
single battery type, depending on the model structure.

@@

The parameterisation of battery models is challenging due to the large
number of parameters that need to be identified compared to the
measurable outputs (\citeproc{ref-Andersson:2022}{Andersson et al.,
2022}; \citeproc{ref-Miguel:2021}{Miguel et al., 2021};
\citeproc{ref-Wang:2022}{Wang et al., 2022}). A complete
parameterisation often requires a stepwise identification of smaller
sets of parameters from a variety of excitations and different data sets
(\citeproc{ref-Chen:2020}{Chen et al., 2020};
\citeproc{ref-Chu:2019}{Chu et al., 2019}; \citeproc{ref-Kirk:2022}{Kirk
et al., 2023}; \citeproc{ref-Lu:2021}{Lu et al., 2021}).

A generic data fitting optimisation problem may be formulated as:
\begin{equation}
\min_{\mathbf{\theta}} ~ \mathcal{L}_{(\hat{\mathbf{y}_i})}(\mathbf{\theta}) ~~~
\textrm{subject to equations (\ref{dynamics})\textrm{-}(\ref{initial_conditions})}
\label{eqn:parameterisation}
\end{equation}

where \(\mathcal{L} : \mathbf{\theta} \mapsto [0,\infty)\) is a cost
function that quantifies the agreement between the model output
\(\mathbf{y}(t)\) and a sequence of observations
\((\hat{\mathbf{y}_i})\) measured at times \(t_i\). Within the
\texttt{PyBOP} framework, the \texttt{FittingProblem} class packages the
model output along with the measured observations, both of which are
then passed to the cost classes for the computation of the specific cost
function. For gradient-based optimisers, the Jacobian of the cost
function with respect to the unknown parameters,
\((\frac{\partial \mathcal{L}}{\partial \theta})\) is computed for step
size and directional information.

>>>>

\usepackage[affil-it]{authblk}
\usepackage{orcidlink}
%% \renewcommand\Authsep{, }
\setlength{\affilsep}{1em}
\author[1%
  %
  ]{Brady Planden%
    \,\orcidlink{0000-0002-1082-9125}\,%
    }
\author[1,2%
  %
  ]{Nicola E. Courtier%
    \,\orcidlink{0000-0002-5714-1096}\,%
    }
\author[3%
  %
  ]{Martin Robinson%
    \,\orcidlink{0000-0002-1572-6782}\,%
    }
\author[4%
  %
  ]{Agriya Khetarpal%
    \,\orcidlink{0000-0002-1112-1786}\,%
    }
\author[2,5%
  %
  ]{Ferran Brosa Planella%
    \,\orcidlink{0000-0001-6363-2812}\,%
    }
\author[1,2%
  %
  \ensuremath\mathparagraph]{David A. Howey%
    \,\orcidlink{0000-0002-0620-3955}\,%
    }

@@

The Python Battery Optimisation and Parameterisation (\texttt{PyBOP})
package provides methods for estimating and optimising battery model
parameters, offering both deterministic and stochastic approaches with
example workflows to assist users. \texttt{PyBOP} enables parameter
identification from data for various battery models, including the
electrochemical and equivalent circuit models provided by the popular
open-source \texttt{PyBaMM} package (\citeproc{ref-Sulzer:2021}{Sulzer
et al., 2021}). Using the same approaches, \texttt{PyBOP} can also be
used for design optimisation under user-defined operating conditions
across a variety of model structures and design goals. \texttt{PyBOP}
facilitates optimisation with a range of methods, with diagnostics for
examining optimiser performance and convergence of the cost and
corresponding parameters. Identified parameters can be used for
prediction, on-line estimation and control, and design optimisation,
accelerating battery research and development.

@@

\texttt{PyBOP} is a Python package providing a user-friendly,
object-oriented interface for optimising battery model parameters.
\texttt{PyBOP} leverages the open-source \texttt{PyBaMM} package
(\citeproc{ref-Sulzer:2021}{Sulzer et al., 2021}) to formulate and solve
battery models. Together, these tools serve a broad audience including
students, engineers, and researchers in academia and industry, enabling
the use of advanced models where previously this was not possible
without specialised knowledge of battery modelling, parameter inference,
and software development. \texttt{PyBOP} emphasises clear and
informative diagnostics and workflows to support users with varying
levels of domain expertise, and provides access to a wide range of
optimisation and sampling algorithms. These are enabled through
interfaces to \texttt{PINTS} (\citeproc{ref-Clerx:2019}{Clerx et al.,
2019}), \texttt{SciPy} (\citeproc{ref-SciPy:2020}{Virtanen et al.,
2020}), and \texttt{PyBOP}'s own implementations of algorithms such as
adaptive moment estimation with weight decay (AdamW)
(\citeproc{ref-Loshchilov:2017}{Loshchilov \& Hutter, 2017}), gradient
descent (\citeproc{ref-Cauchy:1847}{Cauchy \& others, 1847}), and cuckoo
search (\citeproc{ref-Yang:2009}{Yang \& Suash Deb, 2009}).

\texttt{PyBOP} supports the battery parameter exchange (BPX) standard
(\citeproc{ref-BPX:2023}{Korotkin et al., 2023}) for sharing parameter
sets. These are typically costly to obtain due to the specialised
equipment and time required for characterisation experiments, the need
for domain knowledge, and the computational cost of estimation.
\texttt{PyBOP} reduces the requirements for the latter two by providing
fast parameter estimation methods, standardised workflows, and parameter
set interoperability (via BPX).

\texttt{PyBOP} complements other lithium-ion battery modelling packages
built around \texttt{PyBaMM}, such as \texttt{liionpack} for battery
pack simulation (\citeproc{ref-Tranter2022}{Tranter et al., 2022}) and
\texttt{pybamm-eis} for fast numerical computation of the
electrochemical impedance of any battery model. Identified
\texttt{PyBOP} parameters are easily exportable to other packages.

@@

\texttt{PyBOP} has a layered structure enabling the necessary
functionality to compute forward predictions, process results, and run
optimisation and sampling algorithms. The forward model is solved using
the battery modelling software \texttt{PyBaMM}, with construction,
parameterisation, and discretisation managed by \texttt{PyBOP}'s model
interface to \texttt{PyBaMM}. This provides a robust object construction
process with a consistent interface between forward models and
optimisers. Furthermore, identifiability metrics are provided along with
the estimated parameters (through Hessian approximation of the cost
functions around the optimum point in frequentist workflows, and
posterior distributions in Bayesian workflows).

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth,height=\textheight]{figures/PyBOP_components.drawio.png}
\caption{The core \texttt{PyBOP} architecture with base class
interfaces. Each class provides a direct mapping to a step in the
optimisation workflow. \label{fig:classes}}
\end{figure}

\texttt{PyBOP} formulates the inference process into four key classes:
model, problem, cost (or likelihood), and optimiser (or sampler), as
shown in \autoref{fig:classes}. Each of these objects represents a base
class with child classes constructing specialised functionality for
different workflows. The model class constructs a \texttt{PyBaMM}
forward model with a specified set of equations, initial conditions,
spatial discretisation, and numerical solver. By composing
\texttt{PyBaMM} directly into \texttt{PyBOP}, specialised models can be
constructed alongside the standard models that can also be modified for
different inference tasks. One such example is spatial
re-discretisation, which is required when one or more geometric
parameters are being optimised. In this situation, \texttt{PyBOP}
rebuilds the \texttt{PyBaMM} model only when necessary, reducing the
total number of rebuilds, providing improved performance. Alongside
construction of the forward model, \texttt{PyBOP}'s model class provides
methods for obtaining sensitivities from the prediction, enabling
gradient-based optimisation. A forward prediction, along with its
corresponding sensitivities, is provided to the problem class for
processing and exception control. A standardised data structure is then
provided to the cost classes, which then computes a distance, design, or
likelihood-based metric for optimisation. For point-based optimisation,
the optimisers minimise the cost function or the negative log-likelihood
if a likelihood class is provided. Bayesian inference is provided by
sampler classes, which accept the \texttt{LogPosterior} class and sample
from it using \texttt{PINTS}-based Monte Carlo algorithms at the time of
submission. In the typical workflow, the classes in
\autoref{fig:classes} are constructed in sequence, from left to right in
the figure.

In addition to the core architecture, \texttt{PyBOP} provides several
specialised inference and optimisation features. One example is
parameter inference from electrochemical impedance spectroscopy (EIS)
simulations, where PyBOP discretises and linearises the EIS forward
model into a sparse mass matrix form with accompanying
auto-differentiated Jacobian. This is then translated into the frequency
domain, giving a direct solution to compute the input-output impedance.
In this situation, the forward models are constructed within the spatial
re-discretisation workflow, allowing for geometric parameter inference
from EIS simulations and data.

A second specialised feature is that \texttt{PyBOP} builds on the
\texttt{JAX} (\citeproc{ref-jax:2018}{Bradbury et al., 2018}) numerical
solvers used by \texttt{PyBaMM} by providing \texttt{JAX}-based cost
functions for automatic forward model differentiation with respect to
the parameters. This functionality provides a performance improvement
and allows users to harness many other JAX-based inference packages to
optimise cost functions, such as \texttt{Numpyro}
(\citeproc{ref-numpyro:2019}{Phan et al., 2019}), \texttt{BlackJAX}
(\citeproc{ref-blackjax:2024}{Cabezas et al., 2024}), and \texttt{Optax}
(\citeproc{ref-optax:2020}{DeepMind et al., 2020}).

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4458}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2048}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3494}}@{}}
\caption{List of available model, problem and cost (or likelihood)
classes. \label{tab:subclasses}}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Battery Models
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Problem Types
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cost / Likelihood Functions
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Battery Models
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Problem Types
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cost / Likelihood Functions
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Single-particle model (SPM) & Fitting problem & Sum-squared error \\
SPM with electrolyte (SPMe) & Design problem & Root-mean-squared
error \\
Doyle-Fuller-Newman (DFN) & Observer & Minkowski \\
Many-particle model (MPM) & & Sum-of-power \\
Multi-species multi-reaction (MSMR) & & Gaussian log likelihood \\
Weppner Huggins & & Maximum a posteriori \\
Equivalent circuit model (ECM) & & Volumetric energy density \\
& & Gravimetric energy density \\
\end{longtable}

Similarly, the current algorithms available for optimisation are
presented in \autoref{tab:optimisers}. It should be noted that SciPy
minimize includes several gradient and non-gradient methods. From here
on, the point-based parameterisation and design-optimisation tasks will
simply be referred to as optimisation tasks. This simplification can be
justified by comparing \autoref{eqn:parameterisation} and
\autoref{eqn:design}; deterministic parameterisation is just an
optimisation task to minimise a distance-based cost between model output
and measured values.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4359}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2308}}@{}}
\caption{Currently supported optimisers classified by candidate solution
type, including gradient information.
\label{tab:optimisers}}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gradient-based
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Evolutionary
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
(Meta)heuristic
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gradient-based
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Evolutionary
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
(Meta)heuristic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Weight decayed adaptive moment estimation (AdamW) & Covariance matrix
adaptation (CMA-ES) & Particle swarm (PSO) \\
Improved resilient backpropagation (iRProp-) & Exponential natural
(xNES) & Nelder-Mead \\
Gradient descent & Separable natural (sNES) & Cuckoo search \\
SciPy minimize & SciPy differential evolution & \\
\end{longtable}

In addition to deterministic optimisers \autoref{tab:subclasses},
\texttt{PyBOP} also provides Monte Carlo sampling routines to estimate
distributions of parameters within a Bayesian framework. These methods
construct a posterior parameter distribution that can be used to assess
uncertainty and practical identifiability. The individual sampler
classes are currently composed within \texttt{PyBOP} from the
\texttt{PINTS} library, with a base sampler class implemented for
interoperability and direct integration with \texttt{PyBOP}'s model,
problem, and likelihood classes. The currently supported samplers are
listed in \autoref{tab:samplers}.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1624}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2393}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1368}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2051}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2564}}@{}}
\caption{Sampling methods supported by \texttt{PyBOP}, classified
according to the proposed method. \label{tab:samplers}}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gradient-based
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Adaptive
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Slicing
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Evolutionary
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Other
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gradient-based
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Adaptive
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Slicing
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Evolutionary
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Other
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Monomial gamma & Delayed rejection adaptive & Rank shrinking &
Differential evolution & Metropolis random walk \\
No-U-turn & Haario Bardenet & Doubling & & Emcee hammer \\
Hamiltonian & Haario & Stepout & & Metropolis adjusted Langevin \\
Relativistic & Rao Blackwell & & & \\
\end{longtable}

@@

In general, battery models (after spatial discretisation) can be written
in the form of a differential-algebraic system of equations,
\begin{equation}
\frac{\mathrm{d} \mathbf{x}}{\mathrm{d} t} = f(t,\mathbf{x},\mathbf{\theta}),
\label{dynamics}
\end{equation} \begin{equation}
0 = g(t, \mathbf{x}, \mathbf{\theta}),
\label{algebraic}
\end{equation} \begin{equation}
\mathbf{y}(t) = h(t, \mathbf{x}, \mathbf{\theta}),
\label{output}
\end{equation} with initial conditions \begin{equation}
\mathbf{x}(0) = \mathbf{x}_0(\mathbf{\theta}).
\label{initial_conditions}
\end{equation}

@@

Common battery models include various types of equivalent circuit models
(e.g.~the Thévenin model), the Doyle--Fuller--Newman (DFN) model
(\citeproc{ref-Doyle:1993}{Doyle et al., 1993};
\citeproc{ref-Fuller:1994}{Fuller et al., 1994}) based on porous
electrode theory, and its reduced-order variants including the single
particle model (SPM) (\citeproc{ref-Planella:2022}{Brosa Planella et
al., 2022}) and the multi-species multi-reaction (MSMR) model
(\citeproc{ref-Verbrugge:2017}{Verbrugge et al., 2017}). Simplified
models that retain acceptable predictive accuracy at lower computational
cost are widely used, for example in battery management systems, while
physics-based models are required to understand the impact of physical
parameters on performance. This separation of complexity traditionally
results in multiple parameterisations for a single battery type,
depending on the model structure.

@@

The parameterisation of battery models is challenging due to the large
number of parameters that need to be identified compared to the number
of measurable outputs (\citeproc{ref-Andersson:2022}{Andersson et al.,
2022}; \citeproc{ref-Miguel:2021}{Miguel et al., 2021};
\citeproc{ref-Wang:2022}{Wang et al., 2022}). A complete
parameterisation often requires stepwise identification of smaller sets
of parameters from a variety of excitations and different data sets
(\citeproc{ref-Chen:2020}{Chen et al., 2020};
\citeproc{ref-Chu:2019}{Chu et al., 2019}; \citeproc{ref-Kirk:2022}{Kirk
et al., 2023}; \citeproc{ref-Lu:2021}{Lu et al., 2021}). Furthermore,
parameter identifiability can be poor for a given set of excitations and
data sets, requiring improved experimental design in addition to
uncertainty capable identification methods
(\citeproc{ref-Aitio:2020}{Aitio et al., 2020}).

A generic data-fitting optimisation problem may be formulated as:
\begin{equation}
\min_{\mathbf{\theta}} ~ \mathcal{L}_{(\hat{\mathbf{y}_i})}(\mathbf{\theta}) ~~~
\textrm{subject to equations (\ref{dynamics})\textrm{-}(\ref{initial_conditions})}
\label{eqn:parameterisation}
\end{equation}

where \(\mathcal{L} : \mathbf{\theta} \mapsto [0,\infty)\) is a cost
function that quantifies the agreement between the model output
\(\mathbf{y}(t)\) and a sequence of observations
\((\hat{\mathbf{y}_i})\) measured at times \(t_i\). Within the
\texttt{PyBOP} framework, the \texttt{FittingProblem} class packages the
model output along with the measured observations, both of which are
then passed to the cost classes for the computation of the specific cost
function. For gradient-based optimisers, the Jacobian of the cost
function with respect to unknown parameters,
\(\partial \mathcal{L} / \partial \theta\), is computed for step-size
and directional information.
